---
title: "Multiome tutorial - MultiAssayExperiment"
author: "Xiaosai Yao"
email: "yao.xiaosai@gene.com"
output:
  BiocStyle::pdf_document:
    toc: true
    number_section: true
    self_contained: true
    titlecaps: true
package: "epiregulon"
vignette: >
  %\VignetteIndexEntry{Multiome tutorial - MultiAssayExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Introduction
This tutorial walks through the reprogram-seq dataset. This is a dataset generated by infecting LNCaP cells with NKX2-1 and GATA6 to examine the effects of these TFs on AR activity.

# Installation

```{r, results='hide', message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("epiregulon")
library(epiregulon)
```

Alternatively, you could install from github

```{r setup, message=FALSE, eval=FALSE}
devtools::install_github(repo ='xiaosaiyao/epiregulon')

library(epiregulon)

```

# Data preparation

Single cell preprocessing needs to performed by user's favorite methods prior to using Epiregulon. The following components are required: <br>
1. Peak matrix from scATAC-seq <br>
2. Gene expression matrix from either paired or unpaired scRNA-seq. RNA-seq integration needs to be performed for unpaired dataset. <br>
3. Dimensionality reduction matrix from with either single modalities or joint scRNA-seq and scATAC-seq <br>


```{r}
# load the MAE object
library(scMultiome)

mae <- scMultiome::reprogramSeq()

# peak matrix
PeakMatrix <- mae[["PeakMatrix"]]

# expression matrix
GeneExpressionMatrix <- mae[["GeneExpressionMatrix"]]
rownames(GeneExpressionMatrix) <- rowData(GeneExpressionMatrix)$name

# arrange hash_assigment
GeneExpressionMatrix$hash_assignment <- factor(as.character(GeneExpressionMatrix$hash_assignment),
                                               levels = c("HTO10_GATA6_UTR", "HTO2_GATA6_v2", "HTO8_NKX2.1_UTR", "HTO3_NKX2.1_v2", 
                                                          "HTO1_FOXA2_v2", "HTO4_mFOXA1_v2", "HTO6_hFOXA1_UTR", "HTO5_NeonG_v2"))
# dimensional reduction matrix
reducedDimMatrix <- reducedDim(mae[['TileMatrix500']], "LSI_ATAC")

```
Visualize singleCellExperiment by UMAP
```{r}
# transfer UMAP_combined from TileMatrix to GeneExpressionMatrix
reducedDim(GeneExpressionMatrix, "UMAP_Combined") <- reducedDim(mae[['TileMatrix500']], "UMAP_Combined")
scater::plotReducedDim(GeneExpressionMatrix, 
                       dimred = "UMAP_Combined", 
                       text_by = "Clusters", 
                       colour_by = "Clusters")

```

# Quick start

## Retrieve bulk TF ChIP-seq binding sites 

First, we retrieve the information of TF binding sites collected from Cistrome and ENCODE ChIP-seq, which are hosted on Genomitory. Currently, human genomes HG19 and HG38 and mouse mm10 are available. 

```{r getTFMotifInfo}
grl <- getTFMotifInfo(genome = "hg38")
head(grl)
```

## Link ATAC-seq peaks to target genes

Next, we compute peak to gene correlations using a custom algorithm that has similar performance to ArchR's P2G function.

```{r calculateP2G}
set.seed(1010)
p2g <- calculateP2G(peakMatrix = PeakMatrix, 
                    expMatrix = GeneExpressionMatrix, 
                    reducedDim = reducedDimMatrix,
                    exp_assay = "normalizedCounts",
                    peak_assay = "counts")

p2g

```


## Add TF motif binding to peaks

The next step is to add the TF binding information by overlapping regions of the peak matrix with the bulk chip-seq database loaded in 2. The user can supply either an archR project path and this function will retrieve the peak matrix, or a peakMatrix in the form of a Granges object or RangedSummarizedExperiment.

```{r addTFMotifInfo}

overlap <- addTFMotifInfo(grl = grl, p2g = p2g, peakMatrix = PeakMatrix)
head(overlap)
```

## Generate regulons

A long format dataframe, representing the inferred regulons, is then generated. The dataframe consists of three columns:

* tf (transcription factor)
* target gene
* peak to gene correlation between tf and target gene

```{r, warning=FALSE, getRegulon}

regulon <- getRegulon(p2g = p2g, overlap = overlap, aggregate = FALSE)
regulon

```

## Network pruning (highly recommended)

Epiregulon prunes the network by performing tests of independence on the observed number of cells jointly expressing transcription factor (TF), regulatory element (RE) and target gene (TG) vs the expected number of cells if TF/RE and TG are independently expressed. We implement two tests, the binomial test and the chi-square test. In the binomial test, the expected probability is P(TF, RE) * P(TG), and the number of trials is the total number of cells, and the observed successes is the number of cells jointly expressing all three elements. In the chi-square test, the expected probability for having all 3 elements active is also P(TF, RE) * P(TG) and the probability otherwise is 1- P(TF, RE) * P(TG). The observed cell count for the active category is the number of cells jointly expressing all three elements, and the cell count for the inactive category is n - n_triple.

We calculate cluster-specific p-values if users supply cluster labels. This is useful if we are interested in cluster-specific networks. The pruned regulons can then be used to visualize differential networks for transcription factors of interest. See section on differential networks.


```{r network pruning, results = "hide", message = FALSE}

pruned.regulon <- pruneRegulon(expMatrix = GeneExpressionMatrix,
                               exp_assay = "normalizedCounts",
                               peakMatrix = PeakMatrix,
                               peak_assay = "counts",
                               test = "chi.sq",
                               regulon[regulon$tf %in% c("NKX2-1","GATA6","FOXA1","FOXA2", "AR"),],
                               clusters = GeneExpressionMatrix$Clusters,
                               prune_value = "pval",
                               regulon_cutoff = 0.05
                               )

pruned.regulon
```

## Add Weights

While the `pruneRegulon' function provides statistics on the joint occurrence of TF-RE-TG, we would like to further estimate the strength of regulation. Biologically, this can be interpreted as the magnitude of gene expression changes induced by transcription factor activity. Epiregulon estimates the regulatory potential using one of the four measures: 1) correlation between TF and target gene expression, 2) mutual information between the TF and target gene expression, 3) Wilcoxon test statistics of target gene expression in cells jointly expressing all 3 elements vs cells that do not, or 4) log 2 fold difference of target gene expression in cells jointly expressing all 3 elements vs cells that do not. 

Three measures (correlation, Wilcoxon statistics and log 2 fold difference) give both the magnitude and directionality of changes whereas mutational information is always positive. The correlation and mutual information statistics are computed on the grouped pseudobulks by user-supplied cluster labels, whereas the Wilcoxon and log fold change group cells based on the joint expression of TF, RE and TG in each single cell. 

```{r addWeights, results = "hide", warning = FALSE, message = FALSE}

regulon.w <- addWeights(regulon = pruned.regulon,
                        expMatrix  = GeneExpressionMatrix,
                        exp_assay  = "normalizedCounts",
                        peakMatrix = PeakMatrix,
                        peak_assay = "counts",
                        clusters = GeneExpressionMatrix$Clusters,
                        method = "wilcox")

regulon.w

```

## (Optional) Annotate with TF motifs

So far the gene regulatory network was constructed from TF ChIP-seq exclusively. Some users would prefer to further annotate the regulatory elements with the presence of motifs. We provide an option to annotate peaks with motifs from the Cisbp database. If no motifs are present for this particular factor (as in the case of co-factors or chromatin modifiers), we return NA. If motifs are available for a factor and the RE contains a motif, we return 1. If motifs are available and the RE does not contain a motif, we return 0. 

If the user has already performed motif annotation with ArchR, we could also retrieve the results directly. See `?addMotifScore`

```{r addMotifScore}

regulon.w.motif <- addMotifScore(regulon = regulon.w,
                                 peaks = rowRanges(PeakMatrix),
                                 species = "human",
                                 genome = "hg38")

# if desired, set weight to 0 if no motif is found
regulon.w.motif$weight[regulon.w.motif$motif == 0] <- 0
```

## Calculate TF activity 

Finally, the activities for a specific TF in each cell are computed by averaging expressions of target genes linked to the TF weighted by the test statistics of choice, chosen from either correlation, mutual information, Wilcoxon test statistics or log fold change. 
$$y=\frac{1}{n}\sum_{i=1}^{n} x_i * weights_i$$
where $y$ is the activity of a TF for a cell
$n$ is the total number of targets for a TF
$x_i$ is the log count expression of target i where i in {1,2,...,n}
$weights_i$ is the weight of TF and target i

```{r calculateActivity, results = "hide"}
score.combine <- calculateActivity(expMatrix = GeneExpressionMatrix,
                                   regulon = regulon.w, 
                                   mode = "weight", 
                                   method = "weightedMean", 
                                   exp_assay = "normalizedCounts",
                                   normalize = FALSE)

head(score.combine)
```

# Session Info

```{r}
sessionInfo()
```


